FAIR fission track analysis with geochron@home

### 2025-05-09 geochron@home/manuscript/paper.tex
Updated g@h code and manuscript

\runningtitle{FAIR fission track dating with geochron@home}

@@

geochron@home was written in Python and PostgreSQL. Its code is freely
available for inspection and modification, allowing users to set up
their own geochron@home server. Alternatively, users who would like to
upload data to the GaHa, but do not have the facilities to set up
their own server, may use the server at University College London free
of charge. The GaHa accepts image stacks acquired on any type of
digital microscope, and accomodates fission track data (counts and
length measurements) from external fission track analysis suites such
as Fission Track Studio and Track\emph{Flow}.\medskip

We anticipate that the introduction of FAIR workflows will make
fission track data more accurate and more future proof. Storing
fission track data online will benefit future developments in fission
track thermochronology. For example, archival datasets of peer
reviewed fission track counts can be used to train and improve machine
learning algorithms for automated fission track analysis. We hope that
other geochronological methods will follow the fission track
community's lead in FAIR data processing, with benefits to all the
other Earth Science disciplines that depend on geochronological data.
\end{abstract}

@@

Unlike most other geochronometers, which use mass spectrometers to
estimate parent-daughter ratios, fission tracks are observed under an
optical microscope and counted by a human observer.  In recent years,
digital microscopy has moved fission track data acquisition from the
objective lens of a microscope to the computer screen
\citep{gleadow2009, vanranst2019, gleadow2019}. Ongoing developments
in artificial intelligence generate further opportunities to improve
the throughput and accuracy of fission track data
\citep{nachtergaele2020}. But despite the richness of the digital
datasets produced by these novel tools, fission track data are still
reported as small summary tables. These tables require an unnecessary
degree of trust between the `producer' and `consumer' of the
data.\medskip

@@

The geochron@home workflow separates the acquisition of
microscope images from their analysis, providing the flexibility to
accommodate data from different microscope manufacturers, which
typically come with their own instrument control software. The
workflow can be broken down into five steps.

@@

\item Upload the data to the geochron@home platform. geochron@home is
  a Django web-app with a PostgreSQL database. The database can be
  accessed via a Python API and a (more limited) web-based GUI
  interface. Accessing the API requires administrator
  privileges. Administrators can create projects, samples and grains;
  download data; and set the access rights of `ordinary' users.
  Projects can be private or public. The source code and installation
  instructions for geochron@home are freely available over GitHub (see
  the Data Availability statement at the end of this paper). This
  allows fission track users to set up their own
  server. Alternatively, fission track labs can upload their data to
  the UCL server by contacting the corresponding author.

@@

\item Post-processing. The fission track data can either be downloaded
  as a flat data table of counts and area, or as a \texttt{.json} file
  containing the locations of all the coounted
  tracks. geochron@home does not provide any tools to
  post-process these files. They are meant to be passed on to other
  tools such as spreadsheet applications or IsoplotR
  \citep{vermeesch2018c}.

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=12cm]{EDM.jpg}
  \captionof{figure}{Screenshots of raw fission track data in
    geochron@home. GR: an apatite grain in reflected light; GT: the
    same grain in transmitted light; MR: the corresponding mica
    detector in reflected light; MT: the mica detector in transmitted
    light. White rectangles mark the region of interest (ROI), within
    which an analyst has counted fission tracks by marking their etch
    pits (shown in yellow). The raw data for this figure can be viewed
    on the geochron@home archive
    (\url{https://github.com/pvermees/GaHa}).\medskip}
  \label{fig:EDM}
}%\end{figure}

@@

Administrators can define regions of interest (ROI) and count or edit
the fission track coordinates of any grain in a geochron@home
database. They can also assign other users to groups, and give these
groups access to a subset of the projects in the
database. Administrators have fine control over the permissions of the
groups. For example, they can allow the members of a group to count
fission tracks in the predefined ROIs, or to define their own
ROIs. Groups provide a mechanism to compare and combine the results of
multiple analysts of the same sample. Figure~\ref{fig:AvP} illustrates
this with two sets of fission track density estimates (in tracks per
\textmu{m}\textsuperscript{2}) for the same sample of Mount Dromedary
apatite \citep{green1985b}, analysed by two users (PV and AC).\medskip

The two analysts defined their own ROIs, resulting in slightly
different density estimates
(Figure~\ref{fig:AvP}a). Figure~\ref{fig:AvP}b compares the track
density estimates of the two analysts for 25 apatite grains on a
radial plot, using the assumption that the density estimates are
statistically independent from each other. Let $N_{PV}$ and $N_{AC}$
be the number of spontaneous tracks counted by PV and AC over areas
$A_{PV}$ and $A_{AC}$, respectively. Then Figure~\ref{fig:AvP}b
assumes that the ratio of the estimated track densities
$\hat{\rho}_{PV} = N_{PV}/A_{PV}$ and $\hat{\rho}_{AC} =
N_{AC}/A_{AC}$ is associated with the following uncertainty:
\begin{equation}
  \hat{\sigma} =
  \sqrt{\mbox{var}\left(\frac{\hat{\rho}_{PV}}{\hat{\rho}_{AC}}\right)} =
  \frac{\hat{\rho}_{PV}}{\hat{\rho}_{AC}} \, \sqrt{\frac{1}{N_{PV}} +
    \frac{1}{N_{PV}} }
  \label{eq:pessimistic}
\end{equation}

In reality, the uncertainties are strongly correlated due to the
significant overlap between the counting areas chosen by the two
analysts. This overlap causes the density ratios to be underdispersed
with respect to the assigned uncertainties ($\mbox{MSWD}=0.34$ in
Figure~\ref{fig:AvP}b).  Equation~\ref{eq:pessimistic} can be modified
to remove the error correlation under the assumption that the two
analysts counted exactly the same features in the overlapping area
($A_0$):
\begin{equation}
  \hat{\sigma}' =
  \frac{\hat{\rho}_{PV}}{\hat{\rho}_{AC}}
  \sqrt{\frac{A_{PV}+A_{AC}}{N_{PV}+N_{AC}}}\,
  \sqrt{
    \frac{1}{A_{PV}} + \frac{1}{A_{AC}} - \frac{2\,A_0}{A_{PV}\,A_{AC}}
  }
  \label{eq:optimistic}
\end{equation}

If there is no overlap between the two counting areas (i.e., if
$A_0=0$), then $\hat{\sigma}' \approx \hat{\sigma}$. Conversely, if
the counting areas are identical (i.e., if $A_{PV}=A_{AC}=A_0$), and
if the two analysts produce perfectly consistent counts, then
$\hat{\sigma}'=0$.\medskip

Equation~\ref{eq:optimistic} produces a radial plot
(Figure~\ref{fig:AvP}c) that is overdispersed ($\mbox{MSWD}=4.2$) with
respect to $\hat{\sigma}'$. This suggests that the assumption of
perfect consistency is also wrong. Detailed inspection of the counted
features in the virtual microscope reveals that the two analysts are
in perfect agreement for long fission tracks, but that they
occasionally disagree about the shorter features. The ratio of the
pooled track densities indicates that PV undercounts the Mount
Dromedary apatite by 7\% relative to AC.
\medskip

In summary, it could be said that Equation~\ref{eq:pessimistic}
represents an overly `pessimistic' view of analysts' abilities to
reproduce each other's results, whereas Equation~\ref{eq:optimistic}
is too `optimistic'. The truth lies between the two scenarios.  The
existence of `observer bias' is well documented \citep{tamer2025}.  It
is one of the reasons why fission track analysis is often done
relative to age standards: observer bias does not have to be a problem
provided that it is consistent between grains, and between
samples.\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=\linewidth]{AvP.pdf}
  \captionof{figure}{Comparison of fission track data for Mount
    Dromedary apatite by two analysts (PV = Pieter Vermeesch and AC =
    Andrew Carter). a) the track counts and ROIs of both users for
    grain 4648, shown in blue (PV) and red (AC); b) radial plot for
    the ratio of the track densities of both users for all 25 grains
    analysed by the two analysts, using the `pessimistic' scenario of
    Equation~\ref{eq:pessimistic}; c) radial plot of the same data,
    using the `optimistic' error propagation formula of
    Equation~\ref{eq:optimistic}.
    \medskip}
  \label{fig:AvP}
}%\end{figure}

@@

\begin{enumerate}
\item{\citet{guo2025}}: this is an LA-ICP-MS based fission track study
  of detrital apatite from the northeastern Tibetan Plateau. It
  contains image stacks of semitracks in 1146 apatite grains from 16
  different samples.
\item{\citet{tamer2025}}: this is a round-robin study in which digital
  image stacks of 44 apatite crystals were circulated among 14
  different analysts. These analysts used the FastTrack image analysis
  software (which is part of the Fission Track Studio suite) to define
  their own ROIs and count the semitracks and horizontally confined
  fission tracks in them. GaHa presents the results of the round-robin
  experiment as a ${44}\times{14}$ grid of URLs.
\item{This study}: All the raw fission track data used in this article
  are available on GaHa, along with the post-processing software that
  was used to produce the figures. Together, these resources provide
  the reader with all the information needed to fully reproduce our
  results, `from cradle to grave'. To our knowledge, this is the first
  geochronological study to do so.
\end{enumerate}

@@

The annotations in the current tutorial pages were made by an
experienced fission track analyst (Andrew Carter). They provide a
quick and easy mechanism to train novice users in the art of fission
track analysis. At the time of writing, the tutorial pages offer only
a limited degree of interactivity. Users can click on features to read
the annotations. However, future plans include the addition of fully
interactive `quizzes' (see Section~\ref{sec:conclusions}). Despite the
simplicity of the current tutorials, they are already quite effective,
as demonstrated by the crowd-sourcing experiment described in the next
section of this paper.

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=15cm]{radialcrowd.pdf}
  \captionof{figure}{a) box plots of 25 crowd-sourced sets of fission
    track counts for 70 analysts. b) radial plot comparing the paired
    track counts for the first two grains in the dataset (4649 and
    4673). The estimate of user PV is shown in blue. Analyst AC used a
    different ROI and is not shown.\medskip}
  \label{fig:radialcrowd}
}%\end{figure}

@@

A 1984 interlaboratory comparison study \citep{miller1985} showed that
the average of several fission track age estimates is closer to the
known age of mineral standards than the age obtained by any individual
observer. Unfortunately, routine measurement of fission track samples
by multiple analysts is prohibitively expensive in a normal laboratory
environment. geochron@home changes this, by bringing fission
track dating to a proverbial `county fair' of citizen
scientists.\medskip

Figure~\ref{fig:crowdsourcing} presents the results of a
crowd-sourcing experiment that was carried out at UCL as part of an
undergraduate course in isotope geology. 70~students were asked to
create an account on the geochron@home server by selecting a
unique username and password. After spending a few minutes to complete
the compulsory tutorial (Section~\ref{sec:tutorial}), the students
were asked to count fission tracks in randomly assigned image stacks
of Mount Dromedary apatite. Each student was required to analyse at
least 15 grains, using pre-defined ROIs. In a matter of hours, the
students amassed a dataset of 37,331 fission track counts in 34
grains. This large dataset was parsed into separate data files --one
for each student-- which they processed during an assessed practical
exercise.\medskip

The raw counts are shown as a series of box plots in
Figure~\ref{fig:crowdsourcing}a for the same selection of 25 grains
that were analysed by AC and PV in Section~\ref{sec:private}. The box
plots exhibit significant variability between the grains and between
the students. This inter-grain variability is expected. It reflects
the different sizes of the ROIs, and the variable U-content of the
grains. The variability of the track counts for each grain results
from observer bias. This phenomenon was previously discussed in
Section~\ref{sec:private}. However, the bias is much stronger for the
crowd-sourced dataset than it was for the two `expert' analysts of
Section~\ref{sec:private}.\medskip

Figure~\ref{fig:crowdsourcing}b compares the 42 raw fission track
counts of the first two grains on a radial plot.  The pairwise ratios
of the counts for each analyst are more consistent than the raw counts
of Figure~\ref{fig:crowdsourcing}a.  Students who over- or undercount
the first grain, tend to over- or undercount the second grain as
well. Nevertheless, the results are overdispersed with respect to the
counting uncertainties, even when using the `pessimistic' scenario of
Equation~\ref{eq:pessimistic}.  The poor consistency of the results
reflects the very limited training received by the students. But
despite these inconsistencies, the opinion of an `expert' (PV, shown
in blue) falls the middle of the estimates obtained by the citizen
scientists. This suggests that the collective wisdom of the entire
group of students is greater than that of any individual
analyst.\medskip

As metioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:radialcrowd} is a two-dimensional histogram for the
x-y positions of all the track positions generated by the students in
the two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets.\medskip

Figure~\ref{fig:radialcrowd} also shows that some students counted the
tails of the fission tracks rather than their etch pits, despite being
told the opposite in the tutorial. Fixing this issue will require some
improvements to the tutorial pages (Section~\ref{sec:outlook}).

@@

  Once trained on historical data, AI algorithms can be used to count
  fission tracks automatically. Following the model of Fission Track
  Studio, machine learning can be used to reverse the fission track
  counting process. Instead of asking users to count the fission
  tracks in a sample, the software can ask them to check the results
  proposed by an AI algorithm, and to remove any features that are
  \emph{not} fission tracks.\medskip

@@

Another advantage of the geochron@home wokflow is the
separation of image acquisition and image analysis. This separation
reduces the hardware requirements for fission track dating. It opens
up the possibility to share resources. State-of-the-art digital
microscopes are expensive. Using geochron@home, a single
microscope can serve multiple users and make fission track analysis
more affordable.\medskip

The fission track method has always been a test bed for new
geochronological developments.  Because fission track data are
imprecise, the fission track community has sollicited the help of
statisticians and mathematicians to develop its analytical protocols.
Other geochronological communities are still catching up with concepts
and tools such as overdispersion, mixture modelling and radial plots,
which have been commonplace in fission track dating for decades
\citep{vermeesch2019b}. In a similar vein, the subjective nature of
fission track identification has prompted the fission track community
to organise inter-laboratory comparisons and round-robin studies long
before other geochronological communities
\citep{miller1985,tamer2025}.\medskip

With the development of geochron@home, fission tracks is once
again ahead of the pack in terms of FAIR data analysis.
geochron@home currently only stores images and counts. This
is enough to reproduce the results of fission track studies using the
external detector method, but not for LA-ICP-MS based data. FAIR data
processing of LA-ICP-MS data requires a new generation of mass
spectrometer data reduction software. We are currently working on
this. The development of FAIR ICP-MS data pipelines will not only
benefit fission track dating but other chronometers as well, such as
in-situ U--Pb, Rb--Sr and Lu--Hf \citep{vermeesch2025}.\medskip

With the establishment of FAIR data, the era of `fiction tracks' will
be gone for good, and geochronology will be well placed to avoid the
reproducibility problems that have plagued other fields of science.

@@

\authorcontribution{Pieter Vermeesch designed the study, acquired the
  funding, counted fission tracks and wrote the paper. Jiangping He
  created geochron@home. Tim Band expanded geochron@home and wrote the
  accompanying microscope image acquisition software. Andrew Carter
  provided the samples, prepared the training data and counted fission
  tracks. Rex Galbraith derived Equation~\ref{eq:optimistic}.}

>>>>

\runningtitle{FAIR fission track analysis with geochron@home}

@@

geochron@home was written in Python and Javascript. Its code is freely
available for inspection and modification, allowing users to set up
their own geochron@home server. Alternatively, users who would like to
upload data to the archive, but do not have the facilities to set up
their own server, may use the server at University College London free
of charge. The archive accepts image stacks acquired on any type of
digital microscope, and accomodates fission track data (counts and
length measurements) from external fission track analysis suites such
as Fission Track Studio and Track\emph{Flow}.\medskip

We anticipate that the introduction of FAIR workflows will make
fission track data more accurate and more future proof. Storing
fission track data online will benefit future developments in fission
track thermochronology. For example, archival datasets of peer
reviewed fission track counts can be used to train and improve machine
learning algorithms for automated fission track analysis. We hope that
other geochronological methods will follow the fission track
community's lead in FAIR data processing. This would benefit all the
other Earth Science disciplines that depend on geochronological data.
\end{abstract}

@@

Unlike most other geochronometers, which require mass spectrometers to
estimate parent-daughter ratios, fission tracks are observed under an
optical microscope and counted by a human observer.  In recent years,
digital microscopy has moved fission track data acquisition from the
objective lens of a microscope to the computer screen
\citep{gleadow2009, vanranst2019, gleadow2019}. Ongoing developments
in artificial intelligence generate further opportunities to improve
the throughput and accuracy of fission track data
\citep{nachtergaele2020}. But despite the richness of the digital
datasets produced by these novel tools, fission track data are still
reported as small summary tables. These tables require an unnecessary
degree of trust between the `producer' and `consumer' of the
data.\medskip

@@

The geochron@home workflow separates the acquisition of microscope
images from their analysis, providing the flexibility to accommodate
data from different microscope manufacturers. The workflow can be
broken down into five steps.

@@

\item Upload the data to the geochron@home platform. geochron@home is
  a Django web-app featuring a user interface powered by the Leaflet
  library (JavaScript).  The database can be accessed via a Python API
  and a (more limited) web-based GUI. Accessing the API requires
  administrator privileges. Administrators can create projects,
  samples and grains; download data; and set the access rights of
  `ordinary' users.  Projects can be private or public. The source
  code and installation instructions for geochron@home are freely
  available over GitHub (see the Data Availability statement at the
  end of this paper). This allows fission track users to set up their
  own server. Alternatively, fission track labs can upload their data
  to the UCL server by contacting the corresponding author.

@@

\item Post-processing. The fission track data can either be downloaded
  as a flat data table of counts and area, or as a \texttt{.json} file
  containing the locations of all the counted tracks. geochron@home
  does not provide any tools to post-process these files. They are
  meant to be passed on to other tools such as spreadsheet
  applications or IsoplotR \citep{vermeesch2018c}.

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=12cm]{EDM.jpg}
  \captionof{figure}{
    Screenshots of raw fission track data for the external detector
    method in geochron@home. GR: an apatite grain in reflected light;
    GT: the same grain in transmitted light; MR: the corresponding
    mica detector in reflected light; MT: the mica detector in
    transmitted light. White rectangles mark the region of interest
    (ROI), within which an analyst has counted fission tracks by
    marking their etch pits (shown in yellow). The raw data for this
    figure can be viewed on the geochron@home archive
    (\url{https://github.com/pvermees/GaHa}).\medskip}
  \label{fig:EDM}
}%\end{figure}

@@

Administrators can define regions of interest (ROI) and count or edit
the fission track coordinates of any grain in a geochron@home
database. They can also assign other users to groups, and give these
groups access to a subset of the projects in the
database. Administrators have fine control over the permissions of the
groups. For example, they can allow the members of one group to define
their own ROIs, whilst requiring members of another group to count
fission tracks in predefined ROIs.  Groups provide a mechanism to
compare and combine the results of multiple analysts of the same
sample. Figure~\ref{fig:AvP} illustrates this with two sets of fission
track density estimates (in tracks per \textmu{m}\textsuperscript{2})
for the same sample of Mount Dromedary apatite \citep{green1985b},
analysed by two users (PV and AC).\medskip

The two analysts defined their own ROIs, resulting in slightly
different density estimates (Figure~\ref{fig:AvP}a) for 25 apatite
grains.  Let $N_{PV}$ and $N_{AC}$ be the number of spontaneous tracks
counted by PV and AC over areas $A_{PV}$ and $A_{AC}$,
respectively. Then the ratio of the estimated track densities are
given by $\hat{\rho}_{PV} = N_{PV}/A_{PV}$ and $\hat{\rho}_{AC} =
N_{AC}/A_{AC}$. The uncertainties of these density ratios are strongly
correlated due to the significant overlap between the counting areas
chosen by the two analysts:
\begin{equation}
  \frac{\mbox{se}(\hat{\rho}_{PV}/\hat{\rho}_{AC})}{\hat{\rho}_{PV}/\hat{\rho}_{AC}}
  \approx
  \sqrt{
    \frac{1}{N_{PV}} + \frac{1}{N_{AC}} - \frac{2\,N_{0}}{N_{PV}\,N_{AC}}
  }
  \label{eq:NO}
\end{equation}

\noindent where $N_{0}$ is the number of tracks counted by both
observers in the area of overlap ($A_0$, say) between their respective
ROIs.  If there is no overlap between the two counting areas (i.e., if
$A_0=0$), then Equation~\ref{eq:NO} reduces to the conventional
coefficient of variation for the ratio of two independent Poisson
variables \citep{galbraith1990a}.  Conversely, if the counting areas
are identical (i.e., if $A_{PV}=A_{AC}=A_0$), and if the two analysts
produce perfectly consistent counts (i.e. if $N_{PV} = N_{AC} =
N_{0}$), then Equation~\ref{eq:NO} returns a zero value.\medskip

The radial plot of Figure~\ref{fig:AvP}b contains two sets of counts,
with $\sum N_{PV} = 686$ and $\sum N_{AC} = 679$, and a weighted mean
track density ratio $\hat{\rho}_{PV}/\hat{\rho}_{PV} = 0.94$ with
relative standard error 0.013. PV counted 600 tracks in those common
areas $A_0$, of which 549 were also counted by AC. Conversely, AC
counted 622 tracks of which 549 were also counted by PV. The ratio of
the two analysts' track density estimates based just on the common
area is therefore 600/622 = 0.965 with relative standard error 0.016,
calculated from Equation~\ref{eq:NO}. This is close to the weighted
mean ratio of 0.94 (Figure~\ref{fig:AvP}b), and is slightly nearer to
1. It indicates that PV under-counts the Mount Dromedary apatite by
3.5\% relative to AC.  The existence of `observer bias' is well
documented \citep{tamer2025}.  It is one of the reasons why fission
track analysis is often done relative to age standards: observer bias
does not have to be a problem provided that it is consistent between
grains, and between samples.\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=12cm]{AvP.pdf}
  \captionof{figure}{Comparison of fission track data for Mount
    Dromedary apatite by two analysts (PV = Pieter Vermeesch and AC =
    Andrew Carter). a) the track counts and ROIs of both users for
    grain 4648, shown in blue (PV) and red (AC); b) radial plot for
    the ratio of the track densities of both users for all 25 grains
    analysed by the two analysts, using Equation~\ref{eq:NO}.
    \medskip}
  \label{fig:AvP}
}%\end{figure}

@@

\begin{enumerate}
\item{\citet{guo2025}}: this is an LA-ICP-MS based fission track study
  of detrital apatite from the northeastern Tibetan Plateau. It
  contains image stacks of semitracks in 1146 apatite grains from 16
  different samples.
\item{\citet{tamer2025}}: this is a round-robin study in which digital
  image stacks of 44 apatite crystals were circulated among 14
  different analysts. These analysts used the FastTrack image analysis
  software \citep[which is part of the Fission Track Studio
    suite;][]{gleadow2009} to define their own ROIs and count the
  semitracks and horizontally confined fission tracks in them. GaHa
  presents the results of the round-robin experiment as a
  ${44}\times{14}$ grid of URLs.
\item{This study}: All the raw fission track data used in this article
  are available on GaHa, along with the post-processing software that
  was used to produce the figures. Together, these resources provide
  the reader with all the information needed to fully reproduce our
  results, `from cradle to grave'. To our knowledge, this is the first
  geochronological study to do so.
\end{enumerate}

@@

The annotations in the current tutorial pages were made by an
experienced fission track analyst (Andrew Carter). This basic tutorial
provides a quick and easy mechanism to train novice users in the art
of fission track analysis. The tutorial pages are in their infancy,
and offer only a limited degree of interactivity. Users can click on
features to read the annotations. The tutorial will grow with input
from experienced analysts and future plans include the addition of
fully interactive `quizzes' (Section~\ref{sec:conclusions}). Despite
the simplicity of the current tutorials, they are already quite
effective, as demonstrated by the crowd-sourcing experiment described
in the next section of this paper.

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=15cm]{radialcrowd.pdf}
  \captionof{figure}{a) box plots of 25 crowd-sourced sets of fission
    track counts for first-time student 70 analysts using ROIs defined
    by PV. b) radial plot comparing the paired track counts for the
    first two grains in the dataset (4649 and 4673). The estimate of
    user PV is shown in blue. Analyst AC used a different ROI and is
    not shown.\medskip}
  \label{fig:radialcrowd}
}%\end{figure}

@@

A 1984 interlaboratory comparison study \citep{miller1985} showed that
the average of several fission track age estimates is closer to the
known age of mineral standards than the age obtained by any individual
observer. Unfortunately, routine measurement of fission track samples
by multiple analysts is prohibitively expensive in a normal laboratory
environment. geochron@home changes this, by bringing fission track
analysis to a proverbial `county fair' of citizen scientists.\medskip

Figure~\ref{fig:radialcrowd} presents the results of a crowd-sourcing
experiment that was carried out at UCL as part of an undergraduate
course in isotope geology. 70~students were asked to create an account
on the geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains. This
large dataset was parsed into separate data files --one for each
student-- which they processed during an assessed practical
exercise.\medskip

The raw counts are shown as a series of box plots in
Figure~\ref{fig:radialcrowd}a for the same selection of 25 grains that
were analysed by AC and PV in Section~\ref{sec:private}. The box plots
exhibit significant variability between the grains and between the
students. The inter-grain variability is expected. It reflects the
different sizes of the ROIs, and the variable U-content of the
grains. The inter-student variability is caused by observer bias. This
phenomenon was previously discussed in
Section~\ref{sec:private}. However, the bias is much stronger for the
crowd-sourced dataset than it was for the two `expert' analysts of
Section~\ref{sec:private}. Figure~\ref{fig:radialcrowd}b compares the
42 raw fission track counts of the first two grains on a radial plot,
where
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The pairwise ratios of the counts for each analyst are more consistent
than the raw counts of Figure~\ref{fig:crowdsourcing}a.  Students who
over- or undercount the first grain, tend to over- or undercount the
second grain as well. Nevertheless, the results are overdispersed with
respect to the counting uncertainties. The overdispersion is
particularly bad considering that Equation~\ref{eq:N12} ignores the
systematic uncertainty caused by the fact that all analysts counted
exactly the same grains.  The poor consistency of the results reflects
the very limited training received by the students. But despite these
inconsistencies, the opinion of an `expert' (PV, shown in blue) falls
the middle of the estimates obtained by the citizen scientists. This
suggests that the collective wisdom of the entire group of students is
greater than that of any individual analyst.\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:crowdsourcing} is a two-dimensional histogram for the
x-y positions of all the track positions generated by the students in
the two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets.\medskip

Figure~\ref{fig:crowdsourcing} also shows that some students counted
the tails of the fission tracks rather than their etch pits, despite
being told the opposite in the tutorial. Fixing this issue will
require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).

@@

  Once trained on historical data, AI algorithms can be used to count
  fission tracks automatically. Following the model of Fission Track
  Studio \citep{gleadow2009, gleadow2019}, machine learning can be
  used to reverse the fission track counting process. Instead of
  asking users to count the fission tracks in a sample, the software
  can ask them to check the results proposed by an AI algorithm, and
  to remove any features that are \emph{not} fission tracks.\medskip

@@

Another advantage of the geochron@home wokflow is the separation of
image acquisition and image analysis. This separation reduces the
hardware requirements for fission track geochronology. It opens up the
possibility to share resources. State-of-the-art digital microscopes
are expensive. Using geochron@home, a single microscope can serve
multiple users and make fission track analysis more
affordable.\medskip

The fission track method has always been a test bed for new
geochronological developments.  Because fission track data are
imprecise, the fission track community has sollicited the help of
statisticians and mathematicians to develop its analytical protocols.
Other geochronological communities are still catching up with concepts
and tools such as overdispersion, mixture modelling and radial plots,
which have been commonplace in fission track analysis for decades
\citep{vermeesch2019b}. In a similar vein, the subjective nature of
fission track identification has prompted the fission track community
to organise inter-laboratory comparisons and round-robin studies long
before other geochronological communities
\citep{miller1985,tamer2025}.\medskip

With the development of geochron@home, fission track geochronology is
once again ahead of the pack in terms of FAIR data analysis.
geochron@home currently only stores images and counts. This is enough
to reproduce the results of fission track studies using the external
detector method, but not for LA-ICP-MS based data. FAIR data
processing of LA-ICP-MS data requires a new generation of mass
spectrometer data reduction software. We are currently working on this
\citep{vermeesch2025b}. The development of FAIR ICP-MS data pipelines
will not only benefit fission track analysis but other chronometers as
well, such as in-situ U--Pb, Rb--Sr and Lu--Hf.\medskip

With the establishment of FAIR data, geochronology will be well placed
to avoid the reproducibility problems that have plagued other fields
of science.

@@

\authorcontribution{Pieter Vermeesch designed the study, acquired the
  funding, counted fission tracks and wrote the paper. Jiangping He
  created geochron@home. Tim Band expanded geochron@home and wrote the
  accompanying microscope image acquisition software. Andrew Carter
  provided the samples, prepared the training data and counted fission
  tracks. Rex Galbraith derived Equation~\ref{eq:NO}.}
